Competing Bandits: The Perils of Exploration under Competition

### 2018-09-03 paper/content/inverted_u.tex
edits and comments

\documentclass[../competing_bandits.tex]{subfiles}
\begin{document}
\section{Competition vs Better Algorithms: Inverted-U}\label{section:5}
Now we demonstrate the inverted-U relationship between the
incentivized algorithms and the competition levels. In particular, we
investigate the incentivized algorithms under three different regimes:
permanent monopoly, temporary monopoly, and permanent duopoly:
\swdelete{We break indifference between exploration algorithms and
  non-exploration algorithms towards non-exploration algorithms due to
  deployment costs.}  \swdelete{We summarize the relationship between
  the different competition levels and incentivized algorithms as
  follows:} \swcomment{deleted the discussion of tie-breaking since it
  feels out of place.}
\begin{finding}\label{conj:inverted-u}
  \textit{For sufficiently low warm start, the optimal strategies
    \swcomment{for whom??} in the competition game are:
\begin{center}
\textbf{Permanent Monopoly} - $\DG$ \\
\textbf{Temporary Monopoly} - $\TS$ \\
\textbf{Permanent Duopoly} - $\DG$ \\
\end{center}
The conditions on $\TS$ being dominant under the temporary monopoly are that the incumbent is a temporary monopoly for sufficiently many periods.\swcomment{vague here. }
}
\end{finding}

@@

\xhdr{Permanent Duopoly} \swdelete{Permanent duopoly corresponds to the model
described in section 2 where $X = 0$ so that both firms enter the
market simultaneously. What algorithms are firms incentivized to
deploy? We report the average market share taken over the $N$
simulations for a set of exogenous values of $T_0$. Additionally, we
report the mean and median of the effective end of game.}
\swedit{

  Since in this case both firms enter at the same time, they will both
  get an expected market share of $1/2$ whenever they deploy the same
  algorithms.

To see what algorithms are the firms incentivized to deploy, we look at the average market
share taken over the $N$ simulations for different values of warm-up length $T_0$.

}

\swdelete{Since both firms enter at the same time the game is symmetric and the competition game is zero-sum so we only need to compute the expected payoffs of three pairs of strategies. We do not report them in the figures, but when both firms play the same algorithm the expected market share is 50/50. Since our results are from simulations, we interpret algorithm $A$ being better than algorithm $B$ if the resulting average market share of $A$ is statistically greater than $0.5$. Otherwise, the firm is indifferent between $A$ and $B$ with ties broken towards algorithms that are easier to deploy.}

Tables \ref{sim_ht} and \ref{sim_nih} show the results of the simulations. For the most part Conjecture~\ref{conj:inverted-u} is validated where, for the Heavy Tail and Uniform instances, we see that $\DG$ does better than $\TS$ and is a weakly dominant strategy in the game. However, there are two results that are at odds with the conjecture:

\begin{puzzle}\label{puz:dg_ht}
Even for large warm start, $\DG$ is better than $\DEG$ under Heavy Tail \footnote{We demonstrate this holds even for $K = 3$ in the supplementary material}
\end{puzzle}

\begin{puzzle}\label{puz:nih}
Under Needle In Haystack, $\TS$ is weakly dominant across all warm starts
\end{puzzle}

@@

% latex table generated in R 3.4.0 by xtable 1.8-2 package
% Thu Aug 16 13:13:01 2018
\begin{table}[ht]
\centering
\caption{Duopoly Experiment Heavy Tail}
\begin{tabular}{rlll}
  \hline
 & $T_0$ = 20 & $T_0$ = 250 & $T_0$ = 500 \\ 
  \hline
TS vs DG & \makecell{\textbf{0.29} $\pm$0.03\\ eeog \\ avg: 55\\ med: 0} & \makecell{\textbf{0.72} $\pm$0.02\\ eeog \\ avg: 570\\ med: 0} & \makecell{\textbf{0.76} $\pm$0.02\\ eeog \\ avg: 620\\ med: 98.5} \\ 
  TS vs DEG & \makecell{\textbf{0.3} $\pm$0.03\\ eeog \\ avg: 37\\ med: 0} & \makecell{\textbf{0.88} $\pm$0.01\\ eeog \\ avg: 480\\ med: 0} & \makecell{\textbf{0.9} $\pm$0.01\\ eeog \\ avg: 570\\ med: 113.5} \\ 
  DG vs DEG & \makecell{\textbf{0.62} $\pm$0.03\\ eeog \\ avg: 410\\ med: 7} & \makecell{\textbf{0.6} $\pm$0.02\\ eeog \\ avg: 790\\ med: 762} & \makecell{\textbf{0.57} $\pm$0.03\\ eeog \\ avg: 730\\ med: 608} \\ 
   \hline
\end{tabular}
\label{sim_ht}
\end{table}

\begin{table}[ht]
\centering
\caption{Duopoly Experiment Needle In Haystack}
\begin{tabular}{rlll}
  \hline
 & $T_0$ = 20 & $T_0$ = 250 & $T_0$ = 500 \\
  \hline
TS vs DG & \makecell{\textbf{0.64} $\pm$0.03\\ eeog \\ avg: 200\\ med: 27} & \makecell{\textbf{0.6} $\pm$0.03\\ eeog \\ avg: 370\\ med: 0} & \makecell{\textbf{0.64} $\pm$0.03\\ eeog \\ avg: 580\\ med: 121.5} \\
  TS vs DEG & \makecell{\textbf{0.57} $\pm$0.03\\ eeog \\ avg: 150\\ med: 14} & \makecell{\textbf{0.52} $\pm$0.03\\ eeog \\ avg: 460\\ med: 78.5} & \makecell{\textbf{0.56} $\pm$0.02\\ eeog \\ avg: 740\\ med: 627.5} \\
  DG vs DEG & \makecell{\textbf{0.46} $\pm$0.03\\ eeog \\ avg: 340\\ med: 128.5} & \makecell{\textbf{0.42} $\pm$0.02\\ eeog \\ avg: 650\\ med: 408} & \makecell{\textbf{0.42} $\pm$0.02\\ eeog \\ avg: 690\\ med: 466.5} \\
   \hline
\end{tabular}
\label{sim_nih}
\caption*{\tiny{The first line in each cell contains the average market share received by the firm playing Alg 1 (and the market share of Alg 2 is 1 - Alg 1 Market Share) as well as a 95\% confidence band. For example, the cell in the top left indicates that TS gets on average 64\% of the market when played against DG. The next line contain the average and median effective end of game for this set of simulations.}}
\end{table}
% latex table generated in R 3.4.0 by xtable 1.8-2 package
% Thu Aug 16 13:13:00 2018

\xhdr{Temporary Monopoly} We now consider asymmetries in the timing of entry so that one firm enters the market before the other and serves as a monopolist in the periods until the other firm enters. In terms of our model, this corresponds to varying the value of $X$. For this section we report results fixing $T_0 = 20$ and focus on the family of instances where we observed that, for this warm start, $\DG$ was the optimal strategy \footnote{However, we report the results for the Needle In Haystack instances in the supplementary material. The results are unsurprising, where $\TS$ is the dominant strategy for both the entrant and incumbent.}. It seems clear that the temporary monopoly will get a larger market share regardless of what algorithm it plays, but is the incumbent incentivized to commit to $\TS$?

\begin{finding}
\textit{Counterintuitively, allowing one firm to be a temporary monopolist for a sufficiently long time will induce that firm to commit to the better learning algorithm, $\TS$, even if, under the permanent duopoly, they would commit to $DG$}
\end{finding}

In table \ref{ht_incum} we can see that, for $X = 200$, $\TS$ is a dominant strategy for the incumbent and that $\DG$ is the dominant strategy for the entrant on the Heavy Tail instances \footnote{The result for the Uniform instances is the same, though it takes a higher $X$ to induce the incumbent to play $\TS$.}. In the supplementary material we report the same experiment for different values of $X$ and show that our main result that, for sufficiently large $X$, $\TS$ is the dominant strategy for the incumbent is robust across the instances we consider.

% latex table generated in R 3.4.0 by xtable 1.8-2 package
% Wed Aug 15 19:03:20 2018
\begin{table}[ht]
\centering
\caption{Temporary Monopoly Heavy Tail X = 200}
\begin{tabular}{rlll}
\hline
\multicolumn{4}{c}{Incumbent Algorithm}\\
\multirow{12}{0.6in}{\rotatebox{90}{Entrant Algorithm}} \\
  \hline
 & TS & DEG &  DG \\
  \hline
TS & \makecell{\textbf{0.003} $\pm$0.003\\Var:0.002\\ES:100\%} & \makecell{\textbf{0.083} $\pm$0.02\\Var:0.07\\ES:97\%} & \makecell{\textbf{0.17} $\pm$0.02\\Var:0.1\\ES:95\%} \\
  DEG & \makecell{\textbf{0.045} $\pm$0.01\\Var:0.03\\ES:92\%} & \makecell{\textbf{0.25} $\pm$0.02\\Var:0.1\\ES:75\%} & \makecell{\textbf{0.23} $\pm$0.02\\Var:0.1\\ES:78\%} \\
   DG & \makecell{\textbf{0.12} $\pm$0.02\\Var:0.08\\ES:88\%} & \makecell{\textbf{0.36} $\pm$0.03\\Var:0.2\\ES:76\%} & \makecell{\textbf{0.3} $\pm$0.02\\Var:0.1\\ES:64\%} \\
   \hline
\end{tabular}
\label{ht_incum}
\caption*{\tiny{The first line in each cell contains the average market share for the entrant over $N=1000$ simulations as well as a 95\% confidence interval. The second line contains the sample variance of the observed market shares and the third line contains the fraction of simulations that ended up with one firm getting an ``extreme" market share of $> 90\%$. Note that smaller values in the table are better for the incumbent. Market shares are calculated as the fraction of users selecting a particular firm \textit{after} the entrant has already entered (i.e. the free rounds to the incumbent do not count towards the share)}}
\end{table}

>>>>

\documentclass[../competing_bandits.tex]{subfiles}
\begin{document}
\section{Competition vs Better Algorithms: Inverted-U}\label{section:5}
Now we demonstrate the inverted-U relationship between the
incentivized algorithms and the competition levels. In particular, we
investigate the incentivized algorithms under three different regimes:
permanent monopoly, temporary monopoly, and permanent duopoly:
\swdelete{We break indifference between exploration algorithms and
  non-exploration algorithms towards non-exploration algorithms due to
  deployment costs.}  \swdelete{We summarize the relationship between
  the different competition levels and incentivized algorithms as
  follows:} \swcomment{think the tie-breaking should be discussed in the prelim?}
\begin{finding}\label{conj:inverted-u}
  \textit{For sufficiently low warm start, the optimal strategies
    \swcomment{for whom??} in the competition game are:
\begin{center}
{Permanent Monopoly} - $\DG$ \\
{Temporary Monopoly} - $\TS$ \\
{Permanent Duopoly} - $\DG$ \\
\end{center}
The conditions on $\TS$ being dominant under the temporary monopoly are that the incumbent is a temporary monopoly for sufficiently many periods.\swcomment{can we quantify?}
}
\end{finding}

@@

\xhdr{Permanent Duopoly} \swdelete{Permanent duopoly corresponds to
  the model described in section 2 where $X = 0$ so that both firms
  enter the market simultaneously. What algorithms are firms
  incentivized to deploy? We report the average market share taken
  over the $N$ simulations for a set of exogenous values of
  $T_0$. Additionally, we report the mean and median of the effective
  end of game.}  \swedit{Since both firms enter at the same time, they
  will both get an expected market share of $1/2$ whenever they deploy
  the same algorithms. We will mainly investigate the market shares
  when the firms play different algorithms.  In Tables \ref{sim_ht}
  and \ref{sim_nih}, we report the average market share taken over the
  $N$ simulations for different values of warm-up length $T_0$, and
  also the mean and median of the EEOG, under heavy tail and needle in
  haystack instances.

  For the heavy tail and uniform instances,\footnote{We defer the
    table for uniform instances to the appendix.} we see that $\DG$ is
  a weakly dominant strategy for the competition game, and in
  particular weakly dominates $\TS$. This establishes our claim that
  $\DG$ is the incentivized algorithm.  However, there are two results
  that are at odds with the conjecture: }

\swdelete{Since both firms enter at the same time the game is symmetric and the competition game is zero-sum so we only need to compute the expected payoffs of three pairs of strategies. We do not report them in the figures, but when both firms play the same algorithm the expected market share is 50/50. Since our results are from simulations, we interpret algorithm $A$ being better than algorithm $B$ if the resulting average market share of $A$ is statistically greater than $0.5$. Otherwise, the firm is indifferent between $A$ and $B$ with ties broken towards algorithms that are easier to deploy.

Tables \ref{sim_ht} and \ref{sim_nih} show the results of the simulations. For the most part Conjecture~\ref{conj:inverted-u} is validated where, for the Heavy Tail and Uniform instances, we see that $\DG$ does better than $\TS$ and is a weakly dominant strategy in the game. However, there are two results that are at odds with the conjecture:
}

\swcomment{Guy, please check above! for what instances are $\DG$ a weakly dominant strategy again? Also, what is $N$?}

\begin{puzzle}\label{puz:dg_ht}
Even for large warm start, $\DG$ is better than $\DEG$ under heavy tail instances.\footnote{We demonstrate this holds even for $K = 3$ in the supplementary material}
\end{puzzle}

\begin{puzzle}\label{puz:nih}
Under Needle In Haystack, $\TS$ is weakly dominant across all warm starts. 
\end{puzzle}

@@

% latex table generated in R 3.4.0 by xtable 1.8-2 package
% Thu Aug 16 13:13:01 2018
\begin{table}[ht]
\centering
\caption{Duopoly experiment for heavy tail instances}
\begin{tabular}{rlll}
  \hline
 & $T_0$ = 20 & $T_0$ = 250 & $T_0$ = 500 \\ 
  \hline
TS vs DG & \makecell{\textbf{0.29} $\pm$0.03\\ eeog \\ avg: 55\\ med: 0} & \makecell{\textbf{0.72} $\pm$0.02\\ eeog \\ avg: 570\\ med: 0} & \makecell{\textbf{0.76} $\pm$0.02\\ eeog \\ avg: 620\\ med: 98.5} \\ 
  TS vs DEG & \makecell{\textbf{0.3} $\pm$0.03\\ eeog \\ avg: 37\\ med: 0} & \makecell{\textbf{0.88} $\pm$0.01\\ eeog \\ avg: 480\\ med: 0} & \makecell{\textbf{0.9} $\pm$0.01\\ eeog \\ avg: 570\\ med: 113.5} \\ 
  DG vs DEG & \makecell{\textbf{0.62} $\pm$0.03\\ eeog \\ avg: 410\\ med: 7} & \makecell{\textbf{0.6} $\pm$0.02\\ eeog \\ avg: 790\\ med: 762} & \makecell{\textbf{0.57} $\pm$0.03\\ eeog \\ avg: 730\\ med: 608} \\ 
   \hline
\end{tabular}
\label{sim_ht}
\end{table}

\begin{table}[ht]
\centering
\caption{Duopoly experiment for needle in haystack instances}
\begin{tabular}{rlll}
  \hline
 & $T_0$ = 20 & $T_0$ = 250 & $T_0$ = 500 \\
  \hline
TS vs DG & \makecell{\textbf{0.64} $\pm$0.03\\ eeog \\ avg: 200\\ med: 27} & \makecell{\textbf{0.6} $\pm$0.03\\ eeog \\ avg: 370\\ med: 0} & \makecell{\textbf{0.64} $\pm$0.03\\ eeog \\ avg: 580\\ med: 121.5} \\
  TS vs DEG & \makecell{\textbf{0.57} $\pm$0.03\\ eeog \\ avg: 150\\ med: 14} & \makecell{\textbf{0.52} $\pm$0.03\\ eeog \\ avg: 460\\ med: 78.5} & \makecell{\textbf{0.56} $\pm$0.02\\ eeog \\ avg: 740\\ med: 627.5} \\
  DG vs DEG & \makecell{\textbf{0.46} $\pm$0.03\\ eeog \\ avg: 340\\ med: 128.5} & \makecell{\textbf{0.42} $\pm$0.02\\ eeog \\ avg: 650\\ med: 408} & \makecell{\textbf{0.42} $\pm$0.02\\ eeog \\ avg: 690\\ med: 466.5} \\
   \hline
\end{tabular}
\label{sim_nih}
\caption*{{The first line in each cell contains the average market share received by the firm playing Alg 1 (and the market share of Alg 2 is 1 - Alg 1 Market Share) as well as a 95\% confidence band. For example, the cell in the top left indicates that TS gets on average 64\% of the market when played against DG. The next line contain the average and median effective end of game for this set of simulations.}}
\end{table}
% latex table generated in R 3.4.0 by xtable 1.8-2 package
% Thu Aug 16 13:13:00 2018

\swcomment{Guy: could you add lines and make the table 3 $\times$ 3 grid? Now it's hard to read which number belongs to which cell...}

\xhdr{Temporary Monopoly.} Recall that the incumbent firm enters the
market and serves as a monopolist until the entrant firm enters at
round $X$. For this section we report results fixing $T_0 = 20$ and
focus on the family of instances where we observed that, for this warm
start, $\DG$ was the optimal strategy.\footnote{However, we report the
  results for the Needle In Haystack instances in the supplementary
  material. The results are unsurprising, where $\TS$ is the dominant
  strategy for both the entrant and incumbent.}\swcomment{I don't
  understand this sentence. what's the point?} \swdelete{It seems clear that the
  temporary monopoly will get a larger market share regardless of what
  algorithm it plays.}

\begin{finding}
  \textit{If the incumbent firm can stay as a temporary monopolist for
    a sufficiently long time, it will deploy the better learning
    algorithm, $\TS$, even though, under the permanent duopoly, they
    would commit to $\DG$.}
\end{finding}

In Table \ref{ht_incum} we can see that, for $X = 200$, $\TS$ is a dominant strategy for the incumbent and that $\DG$ is the dominant strategy for the entrant on the heavy tail instances.\footnote{The result for the Uniform instances is the same, though it takes a higher $X$ to induce the incumbent to play $\TS$.} In the supplementary material we report the same experiment for different values of $X$ and show that our main result that, for sufficiently large $X$, $\TS$ is the dominant strategy for the incumbent is robust across the instances we consider.

% latex table generated in R 3.4.0 by xtable 1.8-2 package
% Wed Aug 15 19:03:20 2018
\begin{table}[ht]
\centering
\caption{Temporary Monopoly Heavy Tail X = 200}
\begin{tabular}{rlll}
\hline
\multicolumn{4}{c}{Incumbent Algorithm}\\
\multirow{12}{0.6in}{\rotatebox{90}{Entrant Algorithm}} \\
  \hline
 & TS & DEG &  DG \\
  \hline
TS & \makecell{\textbf{0.003} $\pm$0.003\\Var:0.002\\ES:100\%} & \makecell{\textbf{0.083} $\pm$0.02\\Var:0.07\\ES:97\%} & \makecell{\textbf{0.17} $\pm$0.02\\Var:0.1\\ES:95\%} \\
  DEG & \makecell{\textbf{0.045} $\pm$0.01\\Var:0.03\\ES:92\%} & \makecell{\textbf{0.25} $\pm$0.02\\Var:0.1\\ES:75\%} & \makecell{\textbf{0.23} $\pm$0.02\\Var:0.1\\ES:78\%} \\
   DG & \makecell{\textbf{0.12} $\pm$0.02\\Var:0.08\\ES:88\%} & \makecell{\textbf{0.36} $\pm$0.03\\Var:0.2\\ES:76\%} & \makecell{\textbf{0.3} $\pm$0.02\\Var:0.1\\ES:64\%} \\
   \hline
\end{tabular}
\label{ht_incum}
\caption*{{The first line in each cell contains the average market share for the entrant over $N=1000$ simulations as well as a 95\% confidence interval. The second line contains the sample variance of the observed market shares and the third line contains the fraction of simulations that ended up with one firm getting an ``extreme" market share of $> 90\%$. Note that smaller values in the table are better for the incumbent. Market shares are calculated as the fraction of users selecting a particular firm \textit{after} the entrant has already entered (i.e. the free rounds to the incumbent do not count towards the share)}}
\end{table}
